---
output:
  pdf_document: default
  html_document: default
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(bookdown)
library(svglite)
library(kableExtra)
```

# Long-term drought prediction using deep neural networks based on geospatial weather data {#dp}

*Author: Helena Veit*

*Supervisor: Henri Funk*

*Degree: Master*

```{r dp-table-css, echo=FALSE, results='asis'}
if (knitr::is_html_output()) cat('
<style>
.book-body .page-inner table th,
.book-body .page-inner table td { padding-left: 5px; padding-right: 5px; }
.book-body .page-inner table td:first-child { white-space: nowrap; }
.book-body .page-inner .loss-fig p { text-align: right; margin: 0; }
</style>
')
```

```{r dp-colors, echo=FALSE}
col_mse <- "#3C3C3C"; col_hin <- "#286EA0"; col_pin <- "#C6402A"
dphead <- function(x, color) {
  if (knitr::is_latex_output())
    sprintf("\\textbf{\\textcolor[HTML]{%s}{%s}}", sub("^#", "", color), x)
  else
    sprintf("<strong style='color:%s'>%s</strong>", color, x)
}
```

## Introduction

Drought has no single, universally accepted definition, and which one applies depends on the part of the water cycle and the impacts under consideration [@zargarReviewDroughtIndices2011].
Four types are commonly distinguished, better understood as sequential stages of one process than as independent categories (Figure \@ref(fig:drought-types), adapted from @nandgudeDroughtPredictionComprehensive2023): in a simplified manner a precipitation deficit produces a *meteorological* drought, which depletes soil moisture into an *agricultural* drought, lowers streamflow and groundwater into a *hydrological* drought, and finally reaches water-dependent goods as a *socioeconomic* drought.

```{r drought-types, echo=FALSE, fig.cap="The four drought types as a propagating cascade, adapted from @nandgudeDroughtPredictionComprehensive2023.", out.width="100%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/general/drought_types_vector.svg")
```

This work targets *meteorological drought*, the first stage and earliest signal, driven by a precipitation deficit together with high evapotranspiration.
It can be quantified with various indices, of which this work uses the Standardized Precipitation Evapotranspiration Index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].

<!-- SKETCH
- Motivation: why long-lead drought forecasting matters (agriculture, water management, planning).
- Problem: predicting SPEI-1 one year ahead over the Alps, made hard by nonstationarity (rising drought frequency across the record).
- Approach in brief: a ConvLSTM predicting SPEI-1 maps from 36 months of history, with variations in the loss and in how global and static features are used.
- State the three research questions at a conceptual level, one motivating line each:
    - RQ1 (loss): does the loss function govern the trade-off between point accuracy (RMSE) and drought detection?
    - RQ2 (global features): does FiLM conditioning on large-scale climate indices outperform naive channel injection?
    - RQ3 (static features): does a learned or seasonal topography encoder help over naive injection?
- Optional one to two sentence preview of the main finding (loss dominates a Pareto trade-off, no single winner).
- Contribution and where the work slots in.
-->

## Related Work

Beyond simple baselines such as logistic regression and gradient-boosted trees [@marusovLongtermDroughtPrediction2024a], drought prediction has increasingly turned to machine learning and deep learning [@nandgudeDroughtPredictionComprehensive2023]. The drought prediction task poses two intertwined challenges.
It is temporal, since a future drought has to be inferred from a history of past conditions, which calls for models that handle sequential data. 
It is also spatial, since the data are gridded or measured at related locations, so nearby places are not independent. 
How a model treats these two aspects, and how well it handles each, is what mainly separates the approaches below.

One way to handle the temporal side is the LSTM. 
In its basic form it predicts a drought index from a sequence of past features [@wangDroughtPredictionInsights2023]. 
Other designs extend it, for example folding in nearby stations through a temporal-convolution and attention hybrid [@chenSTATLSTMMultivariateSpatiotemporal2025]. 
Either way, these models capture temporal dependencies well but handle space only through the inputs rather than modeling the field jointly.

Other work models space more directly. 
Graph-based approaches treat grid points as nodes and aggregate over neighbors before a recurrent temporal module, for example a GraphSAGE encoder followed by a GRU with self-attention [@yuMultivariateSpatiotemporalModeling2023]. 
Transformer and foundation models bring spatial forecasting over from weather and climate, including EarthFormer [@gaoEarthformerExploringSpaceTime2023], FourCastNet [@pathakFourCastNetGlobalDatadriven2022], and the Vision-Transformer foundation model ClimaX, pretrained on large climate-simulation archives [@nguyenClimaXFoundationModel2023]. 
These are powerful but designed for large-scale, often global data and substantial compute, and their advantage over simpler models is not guaranteed, since a simple linear baseline can match or outperform transformer forecasters on standard time-series benchmarks [@zengAreTransformersEffective2022].

These approaches share some tendencies, at least among the studies considered here.
They mostly forecast at a short lead, typically one month ahead [@wangDroughtPredictionInsights2023; @chenSTATLSTMMultivariateSpatiotemporal2025; @yuMultivariateSpatiotemporalModeling2023],
the exception being a twelve-month horizon reached by treating drought as a binary classification over global regions [@marusovLongtermDroughtPrediction2024a]. 
Their evaluation usually falls on one side of a divide, either a continuous index scored by average error such as RMSE [@wangDroughtPredictionInsights2023; @chenSTATLSTMMultivariateSpatiotemporal2025]
or discrete drought categories scored by typical classification evaluation metrics such as the F1-score [@yuMultivariateSpatiotemporalModeling2023], rather than a continuous field judged for both accuracy and severe-event detection. 
Aditionally, in these studies the loss function and the handling of input features are left at a single approach rather than compared across different approaches.

This work predicts the continuous SPEI as a spatial field at a twelve-month lead over the Alps, evaluates it for both point accuracy and severe-drought detection,
and, above all, studies how the loss and the conditioning on global and static features shape the trade-off between the two.

## Data and Task

### Domain and source

The study region is the Alpine domain at monthly resolution from 1971 to 2024 (648 months).
The data is derived from the ERA5-Land reanalysis, regridded to this grid and provided in preprocessed form.
A fixed mask marks the valid Alpine cells, shown outlined against the surrounding European topography in Figure \@ref(fig:al-topography).

```{r al-topography, echo=FALSE, fig.cap="Topography over Europe. The red outline marks the Alpine study domain used in this work.", out.width="70%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/raw/topography.svg")
```

### Drought definition: SPEI

SPEI extends the precipitation-only Standardized Precipitation Index (SPI) [@mckeeRelationshipDroughtFrequency1993] by using the water balance, so that temperature enters through evapotranspiration, which can better account for rising temperatures under climate change.

The SPEI is built on the water balance,

$$D = P - PET, \qquad \begin{cases} D > 0 & \text{water surplus} \\ D < 0 & \text{water deficit} \end{cases}$$

where $P$ is precipitation and $PET$ is potential evapotranspiration.

SPEI is defined at a chosen timescale $k$: the monthly water balance is summed over a rolling window of the preceding $k$ months, and the accumulated series is fitted to a distribution and standardized.
Common choices are 1, 3, 6, 12, and 24 months.
Short timescales respond to rapid, month-to-month deficits and longer ones to slower, accumulated dryness, so the timescale sets which kind of drought the index reflects.
This work uses SPEI-1, so each value standardizes the water balance of one month expressed in standard deviations from the local mean.
The index is computed as in @funk2026probabilisticdeeplearningdrought, with a three-parameter gamma fit calibrated over 1990 to 2019.

### Variables

The inputs fall into three groups.

*Dynamic spatial* variables vary in space and time: SPEI (the target, also supplied as history), precipitation, near-surface temperature, surface pressure, and the climatic water balance (precipitation minus potential evapotranspiration).
Together these capture the local water supply and demand and the large-scale circulation that drive meteorological drought.
Figure \@ref(fig:spei-snapshots) shows SPEI, the most directly drought-relevant of these, across a sequence of months.

*Dynamic global* variables vary in time only: the North Atlantic Oscillation index [@hurrellNao1995], grouped Mediterranean sea-surface temperatures, and a cyclical month encoding.
These capture large-scale drivers of European moisture and its seasonality, with the fourteen Mediterranean basins grouped into three regional means (western Mediterranean, eastern Mediterranean, Black Sea) [@millotCirculationMediterraneanSea2005; @lionelloMediterraneanClimateOverview2006], shown in Figure \@ref(fig:med-slt).

*Static spatial* variables vary in space only: the topography, since Alpine elevation strongly shapes precipitation and temperature.

```{r spei-snapshots, echo=FALSE, fig.cap="Raw SPEI over the Alpine domain across a sequence of months, illustrating the spatial and temporal variability the model has to capture.", out.width="100%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/raw/al/spei.svg")
```

```{r med-slt, echo=FALSE, fig.cap="Mediterranean SLT, 1971--2024, aggregated into the same three regional groups (western Mediterranean, eastern Mediterranean, Black Sea) used as global scalar model input.", out.width="100%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/raw/slides/mediterranean_slt.svg")
```

### Preprocessing

Cells outside the Alpine mask are filled and excluded from all statistics.
All variables, including SPEI, are standardized to zero mean and unit variance using statistics from the training years and valid cells only, so that every input shares a common scale and the normalization draws on no information beyond the training period.
SPEI, supplied on its own 1990 to 2019 reference, is thereby brought onto the same training-period footing as the other variables.

### Split and nonstationarity

The data are split chronologically to prevent look-ahead leakage: training on 1971 to 2004, validation on 2005 to 2014, and testing on 2015 to 2024.
Because the split is by time rather than at random, the periods differ in climate.
The average severe-drought frequency rises across them, from 0.53 to 0.62 to 1.04 events per cell per year, so the test period is both out of sample and drier than the training period, as shown per cell in Figure \@ref(fig:drought-freq-severe).

```{r drought-freq-severe, echo=FALSE, fig.cap="Per-cell severe-drought (SPEI $\\le -1.5$) frequency by split. The rightward shift from train to test is why the periods are treated as climatically different rather than identically distributed.", out.width="100%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/processed/exploratory/drought_frequency_severe.svg")
```

### Task & Model

The forecasting task is a continuous, per-cell regression of SPEI-1.
Given thirty-six months of history, the model predicts a single SPEI-1 map at a twelve-month lead, the field for the month one year after the input window.
The forecast is direct rather than recursive: the target horizon is predicted in one step instead of iterating month by month, which avoids the error accumulation of recursive forecasting.

The forecasting model is a Convolutional LSTM (ConvLSTM), following the variant of @marusovLongtermDroughtPrediction2024a.
It replaces the matrix products in the gates of a standard LSTM [@hochreiterLongShorttermMemory1997] with convolutions, so the input $x_t$, hidden state $h_t$, and cell state $c_t$ are three-dimensional feature maps rather than vectors, which preserves spatial structure and models spatial and temporal dependencies jointly.
Figure \@ref(fig:convlstm-schematic) shows the resulting cell.

```{r convlstm-schematic, echo=FALSE, fig.cap="ConvLSTM cell by @marusovLongtermDroughtPrediction2024a. Every gate is a 2D convolution over the concatenated input and hidden-state feature maps, so $x_t$, $h_t$, and $c_t$ stay spatial feature maps throughout.", out.width="75%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/architecture/ConvLSTM.svg")
```

Each gate applies a convolution followed by batch normalization, whose affine shift replaces the additive bias,

$$
\begin{aligned}
f_t &= \sigma(\mathrm{BN}(W_f \ast x_t + U_f \ast h_{t-1})), & i_t &= \sigma(\mathrm{BN}(W_i \ast x_t + U_i \ast h_{t-1})), \\
\tilde{c}_t &= \tanh(\mathrm{BN}(W_c \ast x_t + U_c \ast h_{t-1})), & o_t &= \sigma(\mathrm{BN}(W_o \ast x_t + U_o \ast h_{t-1})),
\end{aligned}
$$

and the cell and hidden states are updated as

$$c_t = f_t \odot c_{t-1} + i_t \odot \tilde{c}_t, \qquad h_t = o_t \odot \tanh(c_t),$$

where $\ast$ denotes convolution, $\odot$ the elementwise product, and $\sigma$ the sigmoid.

The final hidden state is mapped to the predicted SPEI field by a bounded regression head,

$$\hat{y}_T = \lambda \tanh(W_{\text{out}} \ast h_T),$$

where the scaled $\tanh$ keeps the output within $[-\lambda, \lambda]$, with $\lambda = 10$.

## Methods

The experiments address the three questions from the introduction: how the loss function shapes the trade-off between point accuracy and drought detection (RQ1), how dynamic global features are best integrated into the ConvLSTM (RQ2), and how static spatial features are best integrated (RQ3).
Each is treated as one axis of variation, defined below.

### RQ1: Loss functions

The loss determines how strongly errors on dry cells are weighted relative to the rest, and so how hard the model is pushed to predict droughts.
Three families are compared, each averaged over valid cells, with $y$ the observed and $\hat{y}$ the predicted (standardized) SPEI.

<div style="display: grid; grid-template-columns: 1fr 400px; align-items: center; gap: 1.5em; border-left: 5px solid #3C3C3C; padding-left: 1em;">
<div style="min-width: 0;">

`r dphead("Mean squared error", col_mse)` is the symmetric baseline,

$$\mathcal{L}_{\text{MSE}} = (\hat{y} - y)^2.$$

</div>
<div class="loss-fig" style="min-width: 0; width: 360px; text-align: right;">
```{r mse-illustration, echo=FALSE, out.width="170px"}
knitr::include_graphics("work/02-drought_prediction/reports/figures/mse.svg")
```
</div>
</div>

<div style="display: grid; grid-template-columns: 1fr 400px; align-items: center; gap: 1.5em; border-left: 5px solid #286EA0; padding-left: 1em;">
<div style="min-width: 0;">

`r dphead("Hinge-weighted MSE", col_hin)` keeps the squared error but up-weights cells on the dry side of normal,

$$\begin{aligned}
\mathcal{L}_{\text{hwMSE}} &= w\,(\hat{y} - y)^2, \\
w &= 1 + \alpha \max(0, -y), \quad \alpha \in \{1, 5\},
\end{aligned}$$

so a given error costs more the drier the observed cell.

</div>
<div class="loss-fig" style="min-width: 0; width: 360px; text-align: right;">
```{r hinge-illustration, echo=FALSE, out.width=c("170px","170px"), fig.show="hold"}
knitr::include_graphics(c(
  "work/02-drought_prediction/reports/figures/hinge_weight.svg",
  "work/02-drought_prediction/reports/figures/hinge_loss.svg"
))
```
</div>
</div>

<div style="display: grid; grid-template-columns: 1fr 400px; align-items: center; gap: 1.5em; border-left: 5px solid #C6402A; padding-left: 1em;">
<div style="min-width: 0;">

`r dphead("Pinball (quantile) loss", col_pin)` at $\tau = 0.20$ is asymmetric,

$$\mathcal{L}_{\text{pin}} = \begin{cases} \tau\,(y - \hat{y}) & \hat{y} \le y, \\ (1-\tau)\,(\hat{y} - y) & \hat{y} > y, \end{cases}$$

penalizing over-prediction, that is predicting too wet and so missing a drought, by a factor $\tfrac{1-\tau}{\tau} = 4$ more than under-prediction, which biases the model toward the dry tail.

</div>
<div class="loss-fig" style="min-width: 0; width: 360px; text-align: right;">
```{r pinball-illustration, echo=FALSE, out.width="170px"}
knitr::include_graphics("work/02-drought_prediction/reports/figures/pinball.svg")
```
</div>
</div>

### RQ2: Dynamic global features
Two ways of feeding the global features (NAO, Mediterranean SST, month) into the ConvLSTM are compared, contrasted in Figure \@ref(fig:naive-vs-film).
The naive baseline (left) broadcasts each global scalar across the spatial grid and appends it as an extra input channel.

```{r naive-vs-film, echo=FALSE, fig.cap="Naive injection (left) broadcasts each global scalar across the grid and appends it as an input channel. FiLM (right) instead uses the global features to produce a per-channel scale $\\gamma_{t,c}$ and shift $\\beta_{t,c}$ applied to the embedding.", out.width="90%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/architecture/Naive_vs_film.svg")
```

FiLM (feature-wise linear modulation) [@perezFiLMVisualReasoning2018] (right) instead conditions the network by applying a per-channel affine transform to the feature maps,

$$\mathrm{FiLM}(F_c) = \gamma_c\, F_c + \beta_c, \qquad \gamma_c = f_c(x_t), \; \beta_c = h_c(x_t),$$

where $F_c$ is the activation of channel $c$ and the scale $\gamma_c$ and shift $\beta_c$ are learned functions $f_c, h_c$ of the global features $x_t$.
Rather than adding channels, FiLM lets the global state reshape the existing ones. Figure \@ref(fig:convlstm-film) shows where this conditioning enters the ConvLSTM cell introduced above.

```{r convlstm-film, echo=FALSE, fig.cap="Where FiLM sits inside the ConvLSTM cell: it conditions the embedded input feature map before the gate convolutions.", out.width="75%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/architecture/ConvLSTM_FiLM.svg")
```

### RQ3: Static spatial features
Three ways of incorporating the static topography are compared, contrasted in Figure \@ref(fig:naive-vs-static-encoder).
The naive baseline (left) appends the raw static field repeateadly at each time step, treating it the same as a dynamic spacial feature.

```{r naive-vs-static-encoder, echo=FALSE, fig.cap="Naive injection (left) appends the raw static field as constant channels. The learned encoder (right) instead passes it through a small CNN -- one shared convolution for \\emph{single}, or four season-selected convolutions for \\emph{seasonal} -- before concatenating the result.", out.width="90%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/architecture/Naive_vs_single_seasonal.svg")
```

The other two learn a seperate static encoding through a small convolutional network,

$$E(S) = \mathrm{ReLU}(W_2 \ast \mathrm{ReLU}(W_1 \ast S)),$$

with static features $S$ and convolutional kernels $W_1, W_2$ (each with batch normalization), whose output is concatenated to the embedding $x_t^{\text{emb}}$ at every timestep.
The *single* variant uses one shared encoder, giving $[\,x_t^{\text{emb}}, E(S)\,]$, while the *seasonal* variant uses four encoders, one per meteorological season, selected by the season $s(m_t)$ of the target month, giving $[\,x_t^{\text{emb}}, E_{s(m_t)}(S)\,]$. Figure \@ref(fig:convlstm-static) shows where the resulting encoder output enters the cell.

```{r convlstm-static, echo=FALSE, fig.cap="Where the static encoder sits inside the ConvLSTM cell: its output is concatenated to the embedded input and hidden state before the gate convolutions.", out.width="75%", fig.align="center"}
knitr::include_graphics("work/02-drought_prediction/figures/architecture/ConvLSTM_static.svg")
```

### Experimental setup

The three axes are crossed into a full grid of $4 \times 2 \times 3 = 24$ configurations: four losses, by two global-feature strategies, by three static-feature strategies.
All configurations share the chronological split above, a single random seed (42), and the same training protocol: the Adam optimizer with batch size 32, up to 50 epochs with early stopping on the validation RMSE, and the learning rate reduced on plateau.

### Evaluation

All metrics are computed on the test period (2015--2024), over the valid cells and test months.
Point accuracy is measured by the RMSE of the predicted SPEI,

$$\mathrm{RMSE} = \sqrt{\frac{1}{T G}\sum_{t=1}^{T}\sum_{g=1}^{G}\left(\hat{y}_{g,t} - y_{g,t}\right)^2},$$

over $T$ test months and $G$ valid cells.

Detection is evaluated at the severe-drought cutoff $c = -1.5$.
With $d_{g,t} = \mathbb{1}\{y_{g,t} \le c\}$ and $\hat{d}_{g,t} = \mathbb{1}\{\hat{y}_{g,t} \le c\}$ the observed and predicted severe-drought indicators at cell $g$ and month $t$, the confusion counts are pooled over all valid cells and test months,

$$TP = \sum_{t=1}^{T}\sum_{g=1}^{G} d_{g,t}\,\hat{d}_{g,t}, \quad FP = \sum_{t,g} (1 - d_{g,t})\,\hat{d}_{g,t}, \quad FN = \sum_{t,g} d_{g,t}\,(1 - \hat{d}_{g,t}),$$

so that detection is scored from a single confusion matrix over the whole space-time test set rather than a per-cell or per-month average.
The true positive rate (recall) and F1 then follow as

$$\mathrm{TPR} = \frac{TP}{TP + FN}, \qquad \mathrm{F1} = \frac{2\,TP}{2\,TP + FP + FN}.$$

## Results

### Baselines

Three reference forecasts are evaluated on the test period, 2015 to 2024: persistence (the last observed SPEI carried forward twelve months), climatology (the training-period cell mean), and a per-cell linear trend extrapolated to the test years [@wilksStatistical2011; @jolliffeStephenson2003].
They divide the two objectives (Table \@ref(tab:baselines)).
The trend gives the lowest baseline RMSE (1.065), just ahead of climatology (1.077), but neither ever predicts a severe drought, so both score zero on detection.
Persistence is the only baseline that detects drought (F1 0.290, TPR 0.281), at a far higher RMSE (1.349).
The trend therefore sets the bar to beat on RMSE, and persistence the bar to beat on detection.

```{r baselines, echo=FALSE}
b <- data.frame(
  Baseline = c("Persistence", "Climatology", "Trend"),
  Forecast = c("last SPEI carried forward 12 months",
               "per-cell training-period mean",
               "per-cell linear-trend extrapolation"),
  RMSE = c("1.349", "1.077", cell_spec("1.065", bold = TRUE)),
  F1   = c(cell_spec("0.290", bold = TRUE), "0.000", "0.000"),
  TPR  = c(cell_spec("0.281", bold = TRUE), "0.000", "0.000"),
  check.names = FALSE)
kbl(b, escape = FALSE, booktabs = TRUE, align = "llccc",
    caption = "Baseline forecasts on the test set (2015--2024).") |>
  kable_styling(latex_options = "hold_position", full_width = FALSE, font_size = 13)
```

### RQ1: Loss functions

Holding both feature groups at naive injection, the naive/naive configuration, the loss alone produces a clear trade-off between accuracy and detection (Table \@ref(tab:loss-results)).
MSE and the $w=1$ hinge loss attain the lowest RMSE (0.965 and 0.962), both beating the trend floor, but MSE's detection collapses to F1 0.115, far below persistence: minimizing squared error pulls predictions toward the mean and away from the dry tail.
Raising the hinge weight to $w=5$ buys detection (F1 0.297) at the cost of RMSE (1.028).
Pinball pushes furthest and is the only loss to beat persistence on both F1 (0.384) and TPR (0.327), but it carries the highest RMSE (1.155), above the trend floor.
Accuracy and detection thus move in opposite directions as the loss shifts from MSE to Pinball.

```{r loss-results, echo=FALSE}
lc <- c("#3C3C3C", "#6EA5C8", "#286EA0", "#C6402A")
loss <- cell_spec(c("MSE", "$\\mathrm{hwMSE}_{w=1}$",
                    "$\\mathrm{hwMSE}_{w=5}$", "$\\mathrm{Pinball}_{\\tau=0.2}$"),
                  color = lc, bold = TRUE)
gap <- cell_spec(c("+0.100", "+0.103", "+0.037", "-0.090"),
                 color = c("#005028", "#005028", "#005028", "#781414"))
d <- data.frame(
  Loss = c(loss, "Persistence"),
  F1   = c("0.115", "0.277", "0.297", cell_spec("0.384", bold = TRUE), "0.290"),
  TPR  = c("0.064", "0.192", "0.208", cell_spec("0.327", bold = TRUE), "0.281"),
  RMSE = c("0.965", cell_spec("0.962", bold = TRUE), "1.028", "1.155", "1.349"),
  Gap  = c(gap, "---"),
  check.names = FALSE)
kbl(d, escape = FALSE, booktabs = TRUE, align = "lcccc",
    col.names = c("Loss", "F1", "TPR", "RMSE", "Trend RMSE $-$ model RMSE"),
    caption = "Loss functions at naive/naive, on the test set.") |>
  kable_styling(latex_options = "hold_position", full_width = FALSE, font_size = 13)
```

### RQ2: Global features

RQ2 varies only the global-feature pathway, comparing naive/naive with naive/FiLM while the static features stay at naive injection (Table \@ref(tab:film-results); values are naive/FiLM minus naive/naive).
The effect is loss-dependent.
Under MSE and the hinge losses FiLM reduces detection, sharply for the lighter losses (F1 falls 0.107 under MSE and 0.183 under the $w=1$ hinge), while RMSE hardly moves.
Only under Pinball does FiLM improve detection, raising F1 by 0.064 and TPR by 0.131.
FiLM therefore shifts the accuracy-detection balance without setting it.

```{r film-results, echo=FALSE}
lc <- c("#3C3C3C", "#6EA5C8", "#286EA0", "#C6402A")
grn <- "#005028"; red <- "#781414"
loss <- cell_spec(c("MSE", "$\\mathrm{hwMSE}_{w=1}$",
                    "$\\mathrm{hwMSE}_{w=5}$", "$\\mathrm{Pinball}_{\\tau=0.2}$"),
                  color = lc, bold = TRUE)
d <- data.frame(
  Loss = loss,
  f1n = c("0.115","0.277","0.297","0.384"),
  f1v = c("0.008","0.094","0.261","0.448"),
  f1d = cell_spec(c("-0.107","-0.183","-0.035","+0.064"), color = c(red,red,red,grn)),
  tpn = c("0.064","0.192","0.208","0.327"),
  tpv = c("0.004","0.051","0.167","0.459"),
  tpd = cell_spec(c("-0.060","-0.142","-0.041","+0.131"), color = c(red,red,red,grn)),
  rmn = c("0.965","0.962","1.028","1.155"),
  rmv = c("0.943","0.948","1.055","1.160"),
  rmd = cell_spec(c("-0.022","-0.014","+0.027","+0.005"), color = c(grn,grn,red,red)),
  check.names = FALSE)
kbl(d, escape = FALSE, booktabs = TRUE, align = c("l", rep("c", 9)),
    col.names = c("Loss","Naive","FiLM","$\\Delta$","Naive","FiLM","$\\Delta$","Naive","FiLM","$\\Delta$"),
    caption = "naive/FiLM minus naive/naive (global-feature pathway).") |>
  add_header_above(c(" " = 1, "F1" = 3, "TPR" = 3, "RMSE" = 3)) |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 13)
```

The pooled numbers in Table \@ref(tab:film-results) hide where FiLM acts.
Stratifying the test cell-months by observed SPEI shows that its benefit is confined to the two drought categories the $-1.5$ detection threshold covers, severe ($-2 <$ SPEI $\le -1.5$) and extreme (SPEI $\le -2$) (Table \@ref(tab:film-stratified-main)).
FiLM lowers RMSE in both categories, and the gain grows with drought severity and as the loss pushes harder toward the tail, reaching $-0.10$ in the extreme category under Pinball.
The near-normal cells, which make up the vast majority of the domain and dominate the pooled RMSE, are left essentially unchanged, so these gains barely register in the aggregate (only $+0.5\%$ under Pinball).
Under MSE, whose objective ignores the tail, the reduction is negligible, $0.01$ or less in both categories.

```{r film-stratified-main, echo=FALSE}
lc <- c("#3C3C3C", "#6EA5C8", "#286EA0", "#C6402A")
grn <- "#005028"; neu <- "#777777"
loss <- cell_spec(c("MSE", "$\\mathrm{hwMSE}_{w=1}$",
                    "$\\mathrm{hwMSE}_{w=5}$", "$\\mathrm{Pinball}_{\\tau=0.2}$"),
                  color = lc, bold = TRUE)
d <- data.frame(
  Loss = loss,
  sn = c("1.162", "0.915", "0.660", "0.519"),
  sf = c("1.162", "0.890", "0.621", "0.475"),
  sd = cell_spec(c("+0.000", "-0.025", "-0.039", "-0.044"), color = c(neu, grn, grn, grn)),
  en = c("1.520", "1.253", "1.008", "0.832"),
  ef = c("1.509", "1.181", "0.946", "0.735"),
  ed = cell_spec(c("-0.010", "-0.072", "-0.062", "-0.097"), color = grn),
  check.names = FALSE)
kbl(d, escape = FALSE, booktabs = TRUE, align = c("l", rep("c", 6)),
    col.names = c("Loss", "Naive", "FiLM", "$\\Delta$", "Naive", "FiLM", "$\\Delta$"),
    caption = "RMSE by drought category, naive/naive vs naive/FiLM. Severe: $-2 <$ SPEI $\\le -1.5$ (18,572 cell-months); extreme: SPEI $\\le -2$ (5,221).") |>
  add_header_above(c(" " = 1, "Severe" = 3, "Extreme" = 3)) |>
  kable_styling(latex_options = "hold_position", full_width = FALSE)
```

### RQ3: Static features

RQ3 varies only the static encoder, comparing naive/naive with single/naive and seasonal/naive while the global features stay naive (Tables \@ref(tab:single-results) and \@ref(tab:seasonal-results)).
Both encoders show the same loss-dependence and act mainly on detection, leaving RMSE broadly stable.
The single encoder improves detection only under Pinball, where it lifts TPR by 0.214 and lowers RMSE slightly, but reduces detection under the hinge losses.
The seasonal encoder also helps most under Pinball (TPR up 0.220) and, unlike the single encoder, under MSE too, yet it degrades sharply under the $w=5$ hinge (F1 down 0.220).
Neither encoder consistently beats the other.

```{r single-results, echo=FALSE}
lc <- c("#3C3C3C", "#6EA5C8", "#286EA0", "#C6402A")
grn <- "#005028"; red <- "#781414"
loss <- cell_spec(c("MSE", "$\\mathrm{hwMSE}_{w=1}$",
                    "$\\mathrm{hwMSE}_{w=5}$", "$\\mathrm{Pinball}_{\\tau=0.2}$"),
                  color = lc, bold = TRUE)
d <- data.frame(
  Loss = loss,
  f1n = c("0.115","0.277","0.297","0.384"),
  f1v = c("0.096","0.110","0.254","0.438"),
  f1d = cell_spec(c("-0.018","-0.167","-0.042","+0.054"), color = c(red,red,red,grn)),
  tpn = c("0.064","0.192","0.208","0.327"),
  tpv = c("0.054","0.061","0.180","0.541"),
  tpd = cell_spec(c("-0.010","-0.131","-0.027","+0.214"), color = c(red,red,red,grn)),
  rmn = c("0.965","0.962","1.028","1.155"),
  rmv = c("0.964","0.965","1.044","1.144"),
  rmd = cell_spec(c("-0.001","+0.004","+0.016","-0.011"), color = c(grn,red,red,grn)),
  check.names = FALSE)
kbl(d, escape = FALSE, booktabs = TRUE, align = c("l", rep("c", 9)),
    col.names = c("Loss","Naive","Single","$\\Delta$","Naive","Single","$\\Delta$","Naive","Single","$\\Delta$"),
    caption = "single/naive minus naive/naive (static encoder).") |>
  add_header_above(c(" " = 1, "F1" = 3, "TPR" = 3, "RMSE" = 3)) |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 13)
```

```{r seasonal-results, echo=FALSE}
lc <- c("#3C3C3C", "#6EA5C8", "#286EA0", "#C6402A")
grn <- "#005028"; red <- "#781414"
loss <- cell_spec(c("MSE", "$\\mathrm{hwMSE}_{w=1}$",
                    "$\\mathrm{hwMSE}_{w=5}$", "$\\mathrm{Pinball}_{\\tau=0.2}$"),
                  color = lc, bold = TRUE)
d <- data.frame(
  Loss = loss,
  f1n = c("0.115","0.277","0.297","0.384"),
  f1v = c("0.165","0.251","0.077","0.427"),
  f1d = cell_spec(c("+0.051","-0.026","-0.220","+0.043"), color = c(grn,red,red,grn)),
  tpn = c("0.064","0.192","0.208","0.327"),
  tpv = c("0.099","0.176","0.041","0.547"),
  tpd = cell_spec(c("+0.035","-0.016","-0.166","+0.220"), color = c(grn,red,red,grn)),
  rmn = c("0.965","0.962","1.028","1.155"),
  rmv = c("0.971","0.964","1.056","1.142"),
  rmd = cell_spec(c("+0.006","+0.003","+0.028","-0.013"), color = c(red,red,red,grn)),
  check.names = FALSE)
kbl(d, escape = FALSE, booktabs = TRUE, align = c("l", rep("c", 9)),
    col.names = c("Loss","Naive","Seasonal","$\\Delta$","Naive","Seasonal","$\\Delta$","Naive","Seasonal","$\\Delta$"),
    caption = "seasonal/naive minus naive/naive (static encoder).") |>
  add_header_above(c(" " = 1, "F1" = 3, "TPR" = 3, "RMSE" = 3)) |>
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 13)
```

### Overall

Across the full grid no single configuration dominates: the 24 models trace a Pareto front between RMSE and detection, shown together with the persistence and climatology baselines (Figure \@ref(fig:frontier)).

The loss sets the region of the front, with MSE at low RMSE and near-zero detection, Pinball at high detection and high RMSE, and the hinge-weighted losses in between, while the global and static variants move models along the front rather than redefining it.
The loss function is thus the dominant control over whether the model favors accuracy or detection.

```{r frontier, echo=FALSE, fig.cap="RMSE against drought detection, as TPR (left) and F1 (right), for all 24 configurations and the persistence and climatology baselines. The vertical line marks the linear-trend RMSE; points to its left beat the trend on RMSE.", out.width="48%", fig.show="hold", fig.align="center"}
knitr::include_graphics(c(
  "work/02-drought_prediction/figures/frontier_tpr_light.svg",
  "work/02-drought_prediction/figures/frontier_f1_light.svg"
))
```

## Discussion

<!-- SKETCH
- Summary of results against the three questions: loss dominates the accuracy/detection trade-off (RQ1); FiLM vs naive for global features, including the extremes (RQ2); learned single vs seasonal static encoders (RQ3).
- Interpretation: a Pareto trade-off, no single winner; all models sit near the trend RMSE floor at the 12-month lead.
- Open caveat: whether Pinball's detection gain is genuine skill or a negative prediction bias crossing the threshold.
- Key limitations: single seed, single region (Alps), single lead time (12 months), single index timescale (SPEI-1); the supplied SPEI calibration window (1990-2019) overlaps the test period; the detection threshold and RMSE are expressed in training-standardized units.
-->

## Conclusion

<!-- SKETCH (very short: concise summary + outlook)
- Recap: at a 12-month lead over the Alps, the loss function governs the trade-off between point accuracy and severe-drought detection, and the feature conditioning shifts it at the margin.
- Outlook: multiple seeds or ensembles, more regions and lead times, resolving the skill-vs-bias question, richer conditioning.
-->

## Appendix

<!-- TODO: citations still missing from the .bib -- Fisher and Rudin 2019 (model
     reliance), Hooker et al. 2021 (occlude vs. permute for correlated inputs),
     Watson 2022 (extremes-focused evaluation). Full numbers and narrative behind
     the prose below: reports/interpretability_report.md. -->

